---
title: 'We Are What We Eat: Getting the Data into R'
author: Lauren
date: '2018-10-03'
slug: we-are-what-we-eat-getting-the-data-into-r
categories: 
  - American Diet
tags:
  - Data cleaning
  - R
  - Excel
  - Web Scraping
  - Rvest
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: 'post'
---

## Getting the USDA Data From Excel Spreadsheets

After reading several articles here, here, and here, I was inspired to investigate the USDA's food
production and consumption data myself.  Most of the articles used food
production for their analysis (this data is provided by USDA in a csv file), but I wanted to instead
focus on consumption which the USDA has calculated by subtracting their estimates for food loss.
For example, food can be lost XXXX.

However, the consumption data only exists in Excel spreadsheets, which I had not yet had to deal with
in R.  This was a learning experience for me and thanks to the great work of Hadley Wickham (yet again)
and his readxl package, this task was made much easier than it otherwise would have been.




```{r message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(rvest)

```

```{r warning=FALSE}

setwd('/Users/laureniannolo/Documents/datasets/American Diets Over Time/data/')
filenames = dir(pattern="*.xls")
for( i in 1:length(filenames) )
{


path <- filenames[i]
path_name <- gsub("^\\.+|\\.[^.]*$", "", path) #get rid of .xls
sheetnames <- excel_sheets(path)

mylist <- lapply(excel_sheets(path), read_xls, path = path, range = "A6:Q52", na = c("","6/","NA","N/A")) #/6 is for the notes about italian cheese

# name the dataframes using the sheetnames

names(mylist) <- sheetnames

#remove the table of contents from the list
mylist <- mylist[names(mylist) != 'TableOfContents']


#list2env(mylist ,.GlobalEnv)


#iterate over all the dataframes in the list and remove columns that are completely NA,
#using 90% of column as NA as a threshold to be safe


NAthres <- 0.9

mylist2 <- lapply(mylist, function(df) {
  cols <- apply(df, 2, function(x) sum(is.na(x)))
  return(df[, cols < (nrow(df) * NAthres)])
})

#list2env(mylist2 ,.GlobalEnv)



#then iterate over all the dataframes in the list and name the columns correctly
#depending on if they are a summary tab or a detail tab
#added in edge cases after first iteration to handle some different formatting of tabs

mylist3 <- lapply(mylist2, function(df) {
  ifelse(ncol(df) == 16, 
    #for detail tabs in all files but fat and juice
    colnames(df)<- c("year","primaryweight_lbyr","lossretail_percent","retail_lbyr","lossconsumer_percent","consumer_lbyr",
                     "consumer_nonedibleloss_percent","consumer_cookingloss_percent","totalloss_percent","percapita_lbyr",
                     "percapita_unitday","percapita_gday","caloriesperunit","unit_equiv","calories_daily","foodpattern_uniteqiv"),
  
  ifelse(ncol(df) == 15, #15 is for fat which is missing foodpattern_uniteqiv for detailed tabs and has daily_fat_g
    colnames(df)<- c("year","primaryweight_lbyr","lossretail_percent","retail_lbyr","lossconsumer_percent","consumer_lbyr",
                     "consumer_nonedibleloss_percent","consumer_cookingloss_percent","totalloss_percent","percapita_lbyr",
                     "percapita_unitday","percapita_gday","caloriesperunit","daily_fat_g","calories_daily"),
  
  ifelse(ncol(df) == 17, #17 is for detail fruit juice and milk which has an added column of gallons per year
    colnames(df)<- c("year","primaryweight_lbyr","lossretail_percent","retail_lbyr","lossconsumer_percent","consumer_lbyr",
                      "consumer_nonedibleloss_percent","consumer_cookingloss_percent","totalloss_percent","percapita_lbyr",
                      "percapita_gallonsyr","percapita_unitday","percapita_gday","caloriesperunit","unit_equiv","calories_daily",
                     "foodpattern_uniteqiv"), 
  
  ifelse(ncol(df) == 11,  #11 is for summary juice and milk which has an added column of gallons per year
    colnames(df)<- c("year","primaryweight_lbyr","retail_lbyr","consumer_lbyr","totalloss_percent","percapita_lbyr",
                     "percapita_gallonsyr","percapita_unitday","percapita_gday","calories_daily","foodpattern_uniteqiv"),
    
  ifelse(ncol(df) == 10,  #10 is for summary fat which is missing foodpattern_uniteqiv and has daily_fat_g
    colnames(df)<- c("year","primaryweight_lbyr","retail_lbyr","consumer_lbyr","totalloss_percent","percapita_lbyr",
                            "percapita_unitday","percapita_gday","daily_fat_g","calories_daily"),
    
    #for summary tabs except fruit juice and fat
    colnames(df)<- c("year","primaryweight_lbyr","retail_lbyr","consumer_lbyr","totalloss_percent","percapita_lbyr",
                     "percapita_unitday","percapita_gday","calories_daily","foodpattern_uniteqiv")
    
  
  )))))
  df
})

#list2env(mylist3 ,.GlobalEnv)

#add three additional columns, one with the name of the dataframe, one with the name of the
#excel spreadsheet, and one with either "summary" or "detail"

mylist4 <- mapply(function(df) {
  "datatype" = ifelse((ncol(df) == 16) || (ncol(df) == 15) || (ncol(df) == 17), "detail","summary")
  df <- cbind(df,datatype) 
  df}, mylist3, SIMPLIFY=F)

mylist4 <- mapply(cbind, mylist4, "foodtype" = names(mylist4), SIMPLIFY=F)
mylist4 <- mapply(cbind, mylist4, "datasource" = path_name, SIMPLIFY=F)

#list2env(mylist4 ,.GlobalEnv)


#add missing columns to each dataframe as blank so that they will have the 
#same amount as the max which is 17 per above


mylist5 <- lapply(mylist4, function(df){
  col_list <- c("year","primaryweight_lbyr","lossretail_percent","retail_lbyr","lossconsumer_percent","consumer_lbyr",
            "consumer_nonedibleloss_percent","consumer_cookingloss_percent","totalloss_percent","percapita_lbyr",
            "percapita_gallonsyr","percapita_unitday","percapita_gday","caloriesperunit","unit_equiv","calories_daily",
            "foodpattern_uniteqiv","daily_fat_g")

  add <-col_list[!col_list%in%names(df)]
  
  if(length(add)!=0) df[add] <- NA
  df
}
)

#list2env(mylist5 ,.GlobalEnv)


#reorder the variables so they are all in the same order and can be rbinded together
mylist5.5 <- lapply(mylist5, function(df){
  
  col_list <- c("year",'datatype','foodtype','datasource',"primaryweight_lbyr","lossretail_percent","retail_lbyr","lossconsumer_percent","consumer_lbyr",
    "consumer_nonedibleloss_percent","consumer_cookingloss_percent","totalloss_percent","percapita_lbyr",
    "percapita_gallonsyr","percapita_unitday","percapita_gday","caloriesperunit","unit_equiv","calories_daily",
    "foodpattern_uniteqiv","daily_fat_g")
  
  df[col_list]
  
  })


#rbind all the individual food dataframes together into one long dataframe for each food type

#if the loop has not yet created the list, create it
if(i == 1){
mylist6 <- list()
}

mylist6[[i]] <- assign(paste('food_data_',path_name,sep=""),do.call("rbind", mylist5.5))


}

```
